A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

Freely generated datatypes are widely used both in functional programs and in
logical specifications. They can be implemented efficiently and enjoy nice
properties that can be exploited in automated reasoners. Although they have
their uses in mathematics, they are especially prominent in computer science
applications.
%
However, because datatype values correspond to ground terms, they can only
represent finite objects. For example, the datatype of natural numbers
constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values such as $\const{S}(\const{S}(\ldots))$, a natural
choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. The verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler JinjaThreads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely on codatatypes to capture infinite
processes. Many proof assistants, including Agda, Coq, Isabelle, Matita, and
PVS, explicitly support them, as do the CoALP dialect of Prolog and the Dafny
specification language.

Codatatypes, like datatypes, are freely generated by their constructors. But
whereas datatype values are finite, for codatatypes infinite constructor terms
also denote values. Intuitively, a codatatype consists of all finite and
infinite ground constructor terms, and only those. For example, the
specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{ES}(\const{ES}(\const{ES}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{EZ} \,\mid\, \const{ES}(\ty{enat}) \,\mid\, \const{EInf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{ES}(\infty) \teq \infty$ hold without any additional
work, because both sides of the equality share the same infinite expansion
$\const{ES}(\const{ES}(\const{ES}(\ldots)))$.

  * single decision procedure for datatypes and codatatypes
  * datatypes are implementation following the lines of Barrett et al.
  * codatatypes were added later, motivated by the use of SMT solvers as
    backends to proof assistants (more specifically, CVC4 to Isabelle/HOL)

  * setting (Section~\ref{sec:theory-of-co-datatypes}):
    * universal formulas
    * many-sorted logic
    * mutually (co)recursive types with constructors, selectors, and
      discriminators

@@

  * decision procedure is described abstractly as a calculus
    (Section~\ref{sec:the-calculus})
    * includes rewriting
    * and inferences

@@

@@

    * point to Barrett et al. for SMT datatypes
      * about their own work, they say: "our focus is on generality and
        efficiency rather than immediacy of implementation"
      * (deal more directly with finite sorts than Barrett et al., Section 6.1)
    * additional ones since then (e.g. strings?)
    * anything about codatatypes?
      * proof assistants like Agda, Coq, etc. have them
      * Dafny, CoALP
      * also a lot of theoretical research, some of which is loosely connected,
        e.g. decision procedure for corecursive functions (Henning in Nijmegen)

@@

>>>>

Freely generated datatypes are ubiquitous in functional programs and
logical specifications. They can be implemented efficiently and enjoy
properties that can be exploited in automated reasoners. They are especially
useful to represent finite data structures in computer science applications but
also arise in mathematical formalizations.
%
%However, because datatype values correspond to finite ground terms, they
%are generally not adequate to represent infinite objects.
%For example, the datatype of natural numbers
%constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
%only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to codatatypes (or coinductive datatypes), the
non-well-founded dual of (inductive) datatypes.
%
Although they are sometimes perceived as esoteric, codatatypes have a role to
play in computer science. For example, the verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler JinjaThreads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all rely crucially on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes, like datatypes, are freely generated by their constructors;
but for codatatypes, infinite constructor terms also denote values
(Section~\ref{sec:theory-of-co-datatypes}). Intuitively, a
codatatype consists of all finite and infinite ground constructor terms, and
only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that aptly models the natural numbers extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Inf}\]
%
the codatatype eliminates one case by unifying the finite and infinite cases.
Equations such as $\const{S}(\infty) \teq \infty$ hold without any additional
work, because both sides of the equality share the same infinite expansion
$\const{S}(\const{S}(\const{S}(\ldots)))$.

Datatypes and codatatypes are an integral part of most popular proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, more and more
automatic theorem provers have been extended with dedicated reasoning facilities
for datatypes. The SMT-LIB~2 \cite{barrett-et-al-2010} syntax, implemented by
most SMT (satisfiability modulo theories) solvers, includes a theory of
datatypes.

In this paper, we introduce a unified decision procedure for both datatypes and
codatatypes. The procedure is described abstractly as a calculus
(Section~\ref{sec:the-calculus}). To our knowledge, this is the first decision
procedure for codatatypes. For the datatype case, it roughly follows the lines
of Barrett et al.\ \cite{barrett-et-al-2007}. Because datatypes and codatatypes
share many of the same properties, it is sensible to consider them together.

%  * codatatypes were added later, motivated by the use of SMT solvers as
%    backends to proof assistants (more specifically, CVC4 to Isabelle/HOL)

%  * setting :
    * universal formulas
    * many-sorted logic
    * mutually (co)recursive types with constructors, selectors, and
      discriminators

@@

@@

\newpage

* SMT-Lib

@@

    * point to Barrett et al. for SMT datatypes
      * about their own work, they say: "our focus is on generality and
        efficiency rather than immediacy of implementation"
      * (deal more directly with finite sorts than Barrett et al., Section 6.1)
    * SPASS-Pirate
    * additional ones since then (e.g. strings?)
    * anything about codatatypes?
      * proof assistants like Agda, Coq, etc. have them
      * Dafny, CoALP
      * also a lot of theoretical research, some of which is loosely connected,
        e.g. decision procedure for corecursive functions (Henning in Nijmegen)

@@

  * presented a decision procedure for datatypes and codatatypes

  * part of our program to reduce the gap between automatic and interactive
    theorem proving
  * and to enrich provers with more high-level features that can be implemented
    efficiently
